---
title: "Flights"
author: "Courtney Hodge"
date: "2024-08-29"
output: html_document
---
```{r}
library(nycflights13)
library(tidyverse)

```

1. Load the nycflights13 package, which contains airline on-time data for all flights departing
NYC in 2013. Also includes useful ‘metadata’ on airlines, airports, weather, and planes.
2. Load the tidyverse package
3. Using the flights data,
• find all flights that were less than 1000 miles (distance)
• Keep only the columns: dep_delay, arr_delay, origin, dest, air_time, and
distance
• Add the Z-score for departure delays
• Convert the departure and arrival delays into hours
• Calculate the average flight speed (in mph)
• order by average flight speed (fastest to slowest)
• return the first 12 rows


```{r}
data("flights")
flights |> 
  filter(
    distance< 1000
  ) |> select(
    dep_delay, arr_delay, origin, dest, air_time, distance
  ) |> mutate(
    dep_delay_z = (dep_delay - mean(dep_delay, na.rm = TRUE)) / sd(dep_delay,na.rm = TRUE)
  ) |> mutate(
    dep_delay_hr = dep_delay/60,
    arr_delay_hr = arr_delay/60,
    across(ends_with("delay"), \(x) c/60, .names = "{.col}_hr")
  )
```

